GPU Acceleration of the Material Point Method

### 2019-02-11 pres/pres.tex
impl. begin

\usepackage{textpos}
\usepackage{csquotes}
\usepackage{listings}
\usepackage{xurl}
\urlstyle{sf}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{textcomp}
\usepackage{caption}
\usepackage{color}
\usepackage{subfigure}
\usepackage{wrapfig}
\usepackage[english,ngerman]{babel}
\usepackage[backend=biber,style=alphabetic]{biblatex}
\usepackage{mathtools}
\usepackage{subdepth}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{svg}
\usepackage{xcolor,colortbl}
\usepackage{media9}

@@

\setbeamerfont{frametitle}{size=\small}
\addbibresource{../mscthesis.bib}
\addbibresource{pres.bib}
\captionsetup{justification=raggedright,singlelinecheck=false}
\DeclareMathOperator*{\argmin}{arg min}
\makeatletter
\def\ext@algorithm{lol}% algorithm captions will be written to the .lol file
% share the list making commands and redefine the heading
\AtBeginDocument{%
  \let\l@algorithm\l@lstlisting
  \let\c@algorithm\c@lstlisting
  \let\thealgorithm\thelstlisting
  \renewcommand{\lstlistlistingname}{Algorithms and source code}%
\makeatother

@@

\end{minipage}
\begin{minipage}{0.4\textwidth}
\begin{tikzpicture}
\datavisualization [school book axes,
		    visualize as line/.list={fx},
		    fx={style={color=blue}}
]

@@

\vfill
\begin{minipage}{0.49\textwidth}
  Heaviside step function has no weak derivate: \\\\
  $y= \left\{ \begin{array} { l l } { 0 , } & { \text { if } t < 0 } \\ { 1 , } & { \text { if } t \geq 0 } \end{array} \right.$
\end{minipage}
\begin{minipage}{0.4\textwidth}

\begin{tikzpicture}

  \datavisualization [school book axes,
		      visualize as line/.list={fx},
		      fx={style={color=blue}}]
data[set=fx] {
  x, y
  -1.5, 0
  0, 0
  0, 1
  1.5, 1
};
\end{tikzpicture}
\end{minipage}
\vfill
Allows for point loads, material discontinuities and more. \cite{bathe2006finite}
\end{frame}
\subsection{Discretization of Space and Time}
\begin{frame}
\frametitle{Discretization of Space and Time}
\textbf{Time discretization} with implicit midpoint scheme:
  \begin{equation}
    \frac { y ^ { n + 1 } - y ^ { n } } { \Delta t } = f ^ { n + \frac{1}{2} } = f \left( t ^ { n } + \frac{\Delta t}{2}, \frac{1}{2} y ^ { n } + \frac{1}{2} y ^ { n + 1 } \right)
  \end{equation}
\begin{itemize}
  \item \textbf{implicit} requires linear system solve $\Rightarrow$ more stable, larger time steps
  \item \textbf{midpoint} as it conserves gov. equations
\end{itemize}

@@

\begin{frame}[fragile]
\textbf{Space Discretization} is done in a Galerkin/FEM fashion with grid based interpolants $w_i$ with limited support. Here dyadic products
\begin{equation}
  w_i(\boldsymbol{x}) =w(\boldsymbol{x}-\boldsymbol{x}_i) = w(\frac{1}{h}(x-x_i))w(\frac{1}{h}(y-y_i)w(\frac{1}{h}(z-z_i))
\end{equation}
of cubic b-splines suffice:
\begin{equation}
w ( x ) = \left\{ \begin{array} { l l } { \frac { 1 } { 2 } | x | ^ { 3 } - | x | ^ { 2 } + \frac { 2 } { 3 } } & { 0 \leqslant | x | < 1 } \\ { \frac { 1 } { 6 } ( 2 - | x | ) ^ { 3 } } & { 1 \leqslant | x | < 2 } \\ { 0 } & { 2 \leqslant | x | } \end{array}. \right
\end{equation}
\end{frame}
\begin{frame}

\begin{tikzpicture}[domain=0:1]
\begin{axis}
\addplot {x*x};
\end{axis}
\end{tikzpicture}

\end{frame}
\begin{frame}
Thus set in:
$$
{}_t\boldsymbol{q}(\boldsymbol{x},t^n) = \sum_i \boldsymbol{e}_iw_i(\boldsymbol{x}),
{}_t\boldsymbol{v}^{n(+1)}(\boldsymbol{x}) = \sum_j \boldsymbol{v}^{n(+1)}_jw_j(\boldsymbol{x}).
$$
Combine it with Numerical Integration where the particles function as quadrature points:
\end{frame}
\section{A MPM Guide on GPGPU Pitfalls and Optimizations}
\subsection{d}

\section{Delving Deeper: Further Opportunities}

>>>>

\usepackage{textpos}
\usepackage{csquotes}
\usepackage{listings}
\usepackage{xurl}
\urlstyle{sf}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{textcomp}
\usepackage{caption}
\usepackage{color}
\usepackage{subfigure}
\usepackage{wrapfig}
\usepackage[english]{babel}
\usepackage[backend=biber,style=alphabetic]{biblatex}
\usepackage{mathtools}
\usepackage{subdepth}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{svg}
\usepackage{xcolor,colortbl}
\usepackage{media9}

@@

\setbeamerfont{frametitle}{size=\small}
\addbibresource{../mscthesis.bib}
\addbibresource{pres.bib}
\captionsetup{justification=raggedright,singlelinecheck=false}

\makeatletter
\AtBeginDocument{%
  \let\l@algorithm\l@lstlisting
  \let\c@algorithm\c@lstlisting
  \let\thealgorithm\thelstlisting
  \renewcommand{\lstlistlistingname}{Algorithms and source code}%
}
\makeatother

@@

\end{minipage}
\begin{minipage}{0.4\textwidth}
\begin{tikzpicture}
  \small
\datavisualization [school book axes,
		    y axis = {attribute = y, label},
		    visualize as line/.list={fx},
		    fx={style={color=blue}}
]

@@

\vfill
\begin{minipage}{0.49\textwidth}
  Heaviside step function has no weak derivate: \\\\
  $s= \left\{ \begin{array} { l l } { 0 , } & { \text { if } t < 0 } \\ { 1 , } & { \text { if } t \geq 0 } \end{array} \right.$
\end{minipage}
\begin{minipage}{0.4\textwidth}
\begin{tikzpicture}
  \small
  \begin{axis}[axis lines = middle, width=5.1cm,height=3.5cm,xtick={-1,0,1},ytick={0,1},ymin=-0.3,ymax=1.25,xmin=-1.8,xmax=1.8,ylabel=$s$]
                \addplot[blue,mark=*,mark options={fill=white},samples at={-2.1,0}] {0};
                \addplot[blue,mark=*,samples at={0,2.1}] {1.0};
            \end{axis}
\end{tikzpicture}
\end{minipage}
\vfill
Allows for point loads, material discontinuities and more. \cite{bathe2006finite}
\end{frame}
\subsection{Discretization of Space and Time}
\begin{frame}
\frametitle{Discretization of Space and Time}
\textbf{Time discretization} with implicit midpoint scheme:
  \begin{equation}
    \frac { y ^ { n + 1 } - y ^ { n } } { \Delta t } = f ^ { n + \frac{1}{2} } = f \left( t ^ { n } + \frac{\Delta t}{2}, \frac{1}{2} y ^ { n } + \frac{1}{2} y ^ { n + 1 } \right)
  \end{equation}
\begin{itemize}
  \item \textbf{implicit} requires linear system solve $\Rightarrow$ more stable, larger time steps
  \item \textbf{midpoint} as it conserves gov. equations
\end{itemize}

@@

\begin{frame}[fragile]
\textbf{Space Discretization} is done in a Galerkin/FEM fashion with grid based interpolants $w_i$ with limited support. Here dyadic products
\begin{equation}
  w_i(\boldsymbol{x}) =w(\boldsymbol{x}-\boldsymbol{x}_i) = w(\frac{1}{h}(x-x_i))w(\frac{1}{h}(y-y_i)w(\frac{1}{h}(z-z_i))
\end{equation}
of cubic b-splines suffice:
\begin{equation}
w ( x ) = \left\{ \begin{array} { l l } { \frac { 1 } { 2 } | x | ^ { 3 } - | x | ^ { 2 } + \frac { 2 } { 3 } } & { 0 \leqslant | x | < 1 } \\ { \frac { 1 } { 6 } ( 2 - | x | ) ^ { 3 } } & { 1 \leqslant | x | < 2 } \\ { 0 } & { 2 \leqslant | x | } \end{array}. \right
\end{equation}
\begin{tikzpicture}[
  declare function={
    func(\x)= and(abs(\x) >= 0 , abs(\x)<1) * (pow(abs(\x),3)/2 - pow(abs(\x),2) +2/3)  +
	      and(abs(\x) >= 1 , abs(\x)<2) * (1/6* pow((2-abs(\x)),3))
   ;
  }
]
\begin{axis}[
  width=10cm,height=4cm,
  axis x line=middle, axis y line=middle,
  ymin=-0.1, ymax=1, ytick={0,1}, ylabel=$w$,
  xmin=-2.5, xmax=2.5, xtick={-2,...,2}, xlabel=$x$,
  domain=-2.5:2.5,samples=101, % added
]

\addplot [blue,thick] {func(x)};
\end{axis}
\end{tikzpicture}
\end{frame}
\begin{frame}
Thus set in:
$$
{}_t\boldsymbol{q}(\boldsymbol{x},t^n) = \sum_i \boldsymbol{e}_iw_i(\boldsymbol{x}),
{}_t\boldsymbol{v}^{n(+1)}(\boldsymbol{x}) = \sum_j \boldsymbol{v}^{n(+1)}_jw_j(\boldsymbol{x}).
$$
Combine it with numerical integration where the particles function as quadrature points \cite{Steffen}:

\begin{minipage}{0.6\textwidth}

$$
g _ { i } = \int _ { \Omega } g ( \boldsymbol{x} ) w _ { i } ( \boldsymbol{x} ) d \boldsymbol{x}
$$
\begin{equation}
  \approx \sum _ { p } g _ { p }  w_i(\boldsymbol{x}_p) V _ { p }.
\end{equation}
due to integration by midpoint rule:
\begin{equation}
\int _ { \Omega } f ( x ) d x \approx \sum _ { i = 1 } ^ { N } f \left( x _ { i } \right) h _ { i }.
\end{equation}

\end{minipage}
\begin{minipage}{0.3\textwidth}
  \begin{tikzpicture}[/pgf/declare function={f=-.1*(x)*(x-8);}]
\pgfplotsset{
    integral segments/.code={\pgfmathsetmacro\integralsegments{#1}},
    integral segments=3,
    integral/.style args={#1:#2}{
        ybar interval,
        domain=#1+((#2-#1)/\integralsegments)/2:#2+((#2-#1)/\integralsegments)/2,
        samples=\integralsegments+1,
        x filter/.code=\pgfmathparse{\pgfmathresult-((#2-#1)/\integralsegments)/2}
    }
}
\begin{axis}[
    ylabel=$f$,
    width =5cm,height=5cm,
    domain=0:6,
    samples=100,
    axis lines=middle
]
\addplot [thick] {f};
\addplot [
    red,
    integral=0:6
] {f};
\end{axis}
\end{tikzpicture}
\end{minipage}
\end{frame}
  \subsection{Velocity Fields: APIC-Transfers}
\begin{frame}
  \frametitle{Velocity Fields: APIC-Transfers}
  \begin{minipage}{0.64\textwidth}
  PIC-transfer
  \begin{enumerate}
    \item $(m\boldsymbol{v})_i^n = \sum_p w_{ip}^nm_p\boldsymbol{v}^n_p$
    \item $\boldsymbol{v}_i^n = \frac{(m\boldsymbol{v})_i^n}{m_i^n}$
    \item $\boldsymbol{v}_{p,{PIC}}^{n+1} = \sum_i w^n_{ip}\boldsymbol{v}_i^{n+1}$
  \end{enumerate}
\vspace*{15}
APIC-transfers add a local velocity field $\boldsymbol{C}^n_p$ around $\boldsymbol{v}^n_p$:
\\\vspace*{-10}\\
$ ( m \boldsymbol{v} ) _ { i } ^ { n } = \sum _ { p } w _ { i p } ^ { n } m _ { p } \left( \boldsymbol{v} _ { p } ^ { n } + \boldsymbol{C} _ { p } ^ { n } ( \boldsymbol{x} _ { i } ^ { n } - \boldsymbol{x} _ { p } ^ { n }))$
\end{minipage}
\begin{minipage}{0.30\textwidth}
\begin{minipage}{1.0\textwidth}
  \small
  \includesvg[width=1.0\textwidth]{media/pic.svg}
  P2G-transfer $\sum_p$
\\
\end{minipage}
\begin{minipage}{1.0\textwidth}
  \small
  \includesvg[width=1.0\textwidth]{media/g2p.svg}
  G2P-transfer $\sum_i$
\end{minipage}
\end{minipage}
\end{frame}
\section{A MPM Guide on GPGPU Pitfalls and Optimizations}
\subsection{Layout of the data: SoA vs. AoS}
\begin{frame}[fragile]
\frametitle{Layout of the data: SoA vs. AoS}

\begin{minipage}{0.31\textwidth}
  \tiny
\begin{minipage}{0.95\textwidth}
  SoA-Layout
\begin{lstlisting}[language={GLSL},style={GL},basicstyle=\tiny]
struct Container{
  vec4 positions[n];
  vec4 velocities[n];
} Particles;
\end{lstlisting}
\end{minipage}
\begin{minipage}{0.95\textwidth}
  AoS-Layout
\begin{lstlisting}[language={GLSL},style={GL},basicstyle=\tiny]
struct Particle {
  vec4 position;
  vec4 velocity;
} Particles[n];
\end{lstlisting}
\end{minipage}

\end{minipage}
\begin{minipage}{0.64\textwidth}
  \tiny
  \includesvg[svgpath =../images/,width=\textwidth]{soa}
  Structures of arrays(SoA)
  \\
  \includesvg[svgpath =../images/,width=\textwidth]{aos}
  Arrays of structures(AoS)

\end{minipage}
\end{frame}
\begin{frame}
\begin{table}[htbp]
  \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
    Layout &  $\Delta t_c (\mu s)$	& Speedup &VRAM     & SM     & L2     & SM Issue Util. \\\hline
    AoS         & 243                   & -       &77.7\%   & 7.3\%  & 30.3\% & 6.8\% \\\hline
    SoA         & \textbf{120}                   & 2.26x   &75.4\%   & \textbf{14.3}\% & 29.4\% & \textbf{14.0\%} \\
    \hline
\end{tabular}
\end{table}

  \end{frame}
\section{Delving Deeper: Further Opportunities}
